# Lab 1: Introduction

```{r, include = FALSE}
source("R/booktem_setup.R")
source("R/my_setup.R")
```

## Factorial ANOVA

Consider the following research situation: A psychologist wants to study if and to what extent different behavior of waiters affect the amount of tip money they get, and whether it matters if the behavior is shown by a waiter or waitress.

The researcher distinguishes the following types of behavior: neutral behavior, drawing a smiley on the bill, or making small talk.

They ran a fully crossed experiment with 3 (behaviors) x 2 (gender: waiter or waitress) = 6 conditions. For each condition they collected data for 10 customers who were helped by a waiter showing neutral behavior, 10 helped by a waiter drawing a smiley on the bill, and so forth.

 


Is the design balanced? `r mcq(c(answer = "Yes", "No"))`

What is/are the independent variable(s) in this experiment? `r mcq(c("Gender", "Tip size", answer = "Behavior")[sample.int(3)])`

Draw the conceptual model of the experiment. Then, check your answer.

`r hide("Answer")`

```{r}
library(tidySEM)
library(ggplot2)
lo <- get_layout("", "Gender", "",
                 "Behavior", "", "Tip size", rows = 2)
edges <- data.frame(from = "Behavior", to = "Tip size")
p <- graph_(layout = lo, edges= edges)
plot(p) + geom_segment(aes(x = p$nodes$x[p$nodes$name == "Gender"], xend =  p$nodes$x[p$nodes$name == "Gender"], y = p$nodes$node_ymin[p$nodes$name == "Gender"], yend = p$nodes$y[p$nodes$name == "Behavior"]), arrow = arrow(length = unit(0.03, "npc"), type = "closed"))
```

As you can see, the dependent variable is Tip money, and the independent variables are Type of behavior and Gender. More specifically. Gender is the moderator, as is expected to influence the relationship between Behavior and Tip money.

`r unhide()`





Now, open the dataset `WaiterBehavior.sav`.

We will run the analysis to find out what the results of the experiment are. We can do so with a factorial ANOVA.

First, create all the dummy variables you need, using syntax. Check your answer below.


`r hide("Answer")`

```{r}
library(tidySEM)
library(ggplot2)
lo <- get_layout("", "Gender", "",
                 "Behavior", "", "Tip size", rows = 2)
edges <- data.frame(from = "Behavior", to = "Tip size")
p <- graph_(layout = lo, edges= edges)
plot(p) + geom_segment(aes(x = p$nodes$x[p$nodes$name == "Gender"], xend =  p$nodes$x[p$nodes$name == "Gender"], y = p$nodes$node_ymin[p$nodes$name == "Gender"], yend = p$nodes$y[p$nodes$name == "Behavior"]), arrow = arrow(length = unit(0.03, "npc"), type = "closed"))
```

As you can see, the dependent variable is Tip money, and the independent variables are Type of behavior and Gender. More specifically. Gender is the moderator, as is expected to influence the relationship between Behavior and Tip money.

`r unhide()`


Proceed as follows:

Navigate to Analyze --> General linear model --> Univariate.
Select Tip as the dependent variable and Behavior and Gender as fixed factors.
Via Options, ask for the Descriptive Statistics.
Paste and run the syntax.
Use the information button to verify whether you generated the correct syntax.

 


Consider the table Descriptives Statistics in the SPSS output.

Take a sheet of paper and copy the table that you can find under the information button.

Then, add the cell means, the marginal means for rows and columns, and the overall mean to the table. In total you are expected to insert 12 values.

Figure 2-1.png

 


Take another look at your table. Compute for each cell (by hand) the expected cell mean based on the main effects only. Include these values in the table as well.

In addition, compute the size of the interaction effects for each cell. Add them to the table too.

 


Your table should like something like this:

Waiter

Waitress

Total

Neutral

4.60

(2.63)

1.97

2.20

(4.17)

-1.97

3.40

(-2.43)

Smiley

5.10

(5.58)

-0.48

7.60

(7.12)

0.48

6.36

(0.52)

Small talk

5.50

(6.98)

-1.48

10.00

(8.52)

1.48

7.75

(1.92)

Total

5.0

(-0.77)

6.60

(0.77)

5.83

 

Inspect the values in the table.

Is there evidence for an interaction in this sample?

Correct answer
  Yes 
  No 
 


Draw (by hand) the profile plot on a piece of paper.

Put the type of behavior on the x-axis and draw separate lines for waiters and waitresses.

 


Your graph should like something like this:

Figure 3.png

Does the graph suggest interaction?

If so, describe the interaction effect (i.e., what can we say about the effect of behavior on amount of tip money for waiters and waitresses?).

General answer comments
The lines are not parallel. Hence, also from the graph we see that there is interaction. The effect of type of behavior on the amount of tip money depends on the gender of the waiter/waitress.

It seems that for waiters the tip money does not depend much on the behavior.

For waitresses the effect is stronger; neutral behavior produces the least amount of tip money, whereas small talk is most beneficial.

 


Consult the table Tests of Between Subjects Effects.

How much of the total variance in Tip money is explained by the whole model (i.e., in total by the main and interaction effects together)? Round your answer to three decimal places.

Correct Answers
0.659 (with margin: 0.005)
 


Is the explained variance of the whole model significant at the 5% level?
Correct answer
  Yes 
  No 
General answer comments
To conclude whether the total model explains a significant part of the variance in the dependent variable, we should look at the F-test for the “Corrected Model”.

The p-value for this test is <.001, so we can conclude that the model is significant when using α=.05.

 


We can also calculate effect sizes η2 or partial η2, for the two factors and the interaction effect separately.

For instance, η2A tells us how much variance Factor A explains in tipping.

And, partial η2A tells us how much variance Factor A explains in tipping controlled for the other predictors.

Consider the design we used. Should we interpret η2 or partial η2?

Correct answer
  η2
  Partial η2
General answer comments
In balanced designs we prefer to use  and in unbalanced designs partial η2.

Since this is a balanced design, we should use η2.

 


What is η2A for the factor Behavior? Round your answer to 3 decimal places.
Correct Answers
0.362 (with margin: 0.005)
 


In SPSS we can request partial η2. Even though in this situation it is more appropriate to use η2, let’s try to request partial η2 once for practice.

Navigate to Analyze --> General linear model --> Univariate.
Select Tip as the dependent variable and Behavior and Gender as fixed factors.
Via Options ask for Estimation of effect size.
Paste and run the syntax.
What is the value of partial η2 for the factor Behavior?

Correct Answers
0.515 (with margin: 0.005)
 


Is the interaction effect significant at the 5% level?
Correct answer
  Yes 
  No 
General answer comments
Yes, the interaction effect is significant at the 5% significance level, F(2,54) = 18.315, p < .001.
 


What is the substantive interpretation of this result and what would be the next step (just mention the step)?
General answer comments
This means that the effect of Type of behavior on Tip money depends on the Gender of the waiter/waitress. Hence, we should do a simple effects analysis.
 


We have enough evidence to conclude that in the population, Type of behavior, Gender and the interaction effect explain a significant amount of variance in the amount of tip (based on the Corrected Model significance test). Moreover, we have enough evidence to conclude that the effect of Type of Behavior on the amount of tip depends on the third variable Gender (based on the significance test for the interaction). This means that behavior of waitresses influences tip behavior in a different way than the behavior of waiters!

Our next step would be simple effects analysis!

 


In a previous step you were asked to draw the profile plot by hand, but you can also get the plots in SPSS.

You can do so by clicking on the Plots tab in the ANOVA menu (i.e., via Analyze --> General linear model --> ANOVA), or by changing the syntax as follows:

    UNIANOVA Tip BY Behavior Gender

       /METHOD=SSTYPE(3)

       /INTERCEPT=INCLUDE

       /PLOT=PROFILE(Behavior*Gender)

       /PRINT=HOMOGENEITY DESCRIPTIVE

       /CRITERIA=ALPHA(.05)

       /DESIGN=Behavior Gender Behavior*Gender.

Notice that SPSS puts the first variable on the x-axis (here behavior) and prints separate lines for the second factor (here gender).

Run the syntax and inspect the graph.

 


Change the syntax such that SPSS provides the profile plots with gender on the x-axis and separate lines for behavior. Compare the two profile plots you.

Is there a remarkable difference between the figures?

General answer comments
If you compare the two plots, you see that lines do cross in the first graph, but not in the second graph.

From the first graph, we can conclude that the strength of the effect of the amount of tip money is dependent on the gender of the waiter/waitress.

From the second graph, we can infer that regardless of whether diner was served by a waiter or waitress, the neutral condition on average invokes the least amount of tip money, followed by smileys, and the most profitable is making small talk.

This means that the ordering of the average amount of tip money by type of behavior is the same for waiter and waitress, but the size of the effect differs. Note that, even though we cannot talk about the size of the main effect, we can say something about the effect of behavior in an ordinal way; that is, making small talk is most profitable, neutral least profitable, and smileys in between, but the amount by which something is profitable depends on whether the person who served the food was a waiter or a waitress.

So, it may help to look at the profile plots in different ways!

## Simple Effects Analysis

We have seen that the interaction between Gender and Behavior is significant, so the next step is a simple effects analysis. In this case, we will look at the simple effect of behavior for different levels of Gender. In other words, we will look into the extent to which Behavior affects tip money for waiters and waitresses separately. To do so we have to add something to our syntax:

 

    UNIANOVA Tip BY Behavior Gender

     /METHOD=SSTYPE(3)

     /INTERCEPT=INCLUDE

     /PLOT=PROFILE(Behavior*Gender)

     /PRINT=HOMOGENEITY DESCRIPTIVE

     /CRITERIA=ALPHA(.05)

     /EMMEANS = TABLES(Behavior*Gender) COMPARE(Behavior)

     /DESIGN=Behavior Gender Behavior*Gender.

 

Copy and run this syntax.

 


Consult the table Univariate Tests. This table displays the results of the simple effects tests.

Does the Behavior of waiters have an influence on the amount of tip money people give?

Write down your conclusion AND the corresponding F-statistic, degrees of freedom, and p-value.

General answer comments
We see that the effect of behavior is not significant among waiters. Hence, for waiters we don’t have evidence that the behavior has an effect on average tips received, F(2,52) = 0.591, p = .557.
 


Again, consult the table Univariate Tests. This table gives the results of the simple effects tests.

Does the type of behavior of waitresses have an influence on the tip people give?

Write down your conclusion AND the corresponding F-statistic, degrees of freedom, and p-value.

General answer comments
We do see a significant effect of behavior on average tip money for waitresses, F(2,52) = 46.385, p < .001. Hence, we have convincing evidence that the type of behavior by waitresses affects the average amount of tips. Results suggest that in order to have high tips, waitresses best can make small talk.
 


Inspect the table Pairwise Comparisons. The three experimental conditions are compared in a pairwise manner, split over the factor Gender.

For which pair of groups do the means differ significantly from one another (at the 5% level)?

Correct answer
  Waitress: Neutral-Smiley 
Correct answer
  Waitress: Neutral-Small talk 
Correct answer
  Waitress: Small talk-Smiley 
  Waiter: Neutral-Smiley 
  Waiter: Neutral-Small talk 
  Waiter: Small talk-Smiley 
 


The p-values in the pairwise comparison table are Bonferroni adjusted.

Why do we need to make Bonferroni adjustments?

General answer comments
When doing multiple tests on one sample, like doing these pairwise comparisons, the level of risk of a Type I error increases. To correct for this, we should use an adjusted alpha-level, such as the Bonferroni correction.
 


Does it make sense to adjust your behavior as a waiter/waitress if you want to increase your tip?
Correct answer
  For waitresses behavior does affect amount of tip money received, but for waiters it does not. 
  For waiters behavior does affect amount of tip money received, but for waitresses it does not. 
  For both waiters and waitresses behavior does affect amount of tip money received. 
  For both waiters and waitresses behavior does not affect amount of tip money received. 
General answer comments
For waiters, behavior does not significantly affect the amount tip money received.

For waitresses, the amount of tip money received is dependent on the behavior she shows towards the quests.

Hence, it only makes sense for waitresses to adjust their behavior to increase tip money; smiling works well, but making small talk would result in the biggest increase in tip money received.

 


We finished our simple effects analysis!

Since the interaction effect was significant, we did not need to carry out the “main-effects-only” model. If, however, we would have found the interaction effect to not be significant, we would have had to carry out this main-effects-only model.

For illustration purposes let see how to do this….

 


First, we will create a new syntax via the menu.

Proceed as follows:

Navigate to Analyze --> General linear model --> Univariate.
Now click on [Model…]. SPSS opens a new window.
You see that, by default, SPSS takes the full factorial model. Now we build our own (customized) model. Click on [Custom].
Choose “main effects” as the building term (in the middle). Then select the two factors.
Paste and run the syntax.
If done correctly, the table Test of Between-Subject Effects should not include the interaction effect anymore.

 


How much of the total variance is explained by the main effects only? Round your number to three decimal places.
Correct Answers
0.427 (with margin: 0.005)
General answer comments
The proportion of explained variance in tip money by the main effects only is .427.

This means that of all variance in tips, 42.7% is explained by the main effects only (which is much less than the 65.9% in the full model).

 


Consult the profile plot (if you don’t have the plot, add /PLOT=PROFILE(Behavior*Gender) to your syntax).

We now observe two parallel lines! Actually, this plot is obtained under the main effects only model and represents the estimated means under the main effects only. Take a look at the table you constructed in the first assignment and verify that the profile plots indeed represent expected cell means from the main effects only.

Remark: here we look at the plots just to see how it works in SPSS, but because the interaction is significant, the main effects do not adequately represent the effects (!).

 


In the lecture(s) on ANOVA, we looked at the assumption of homogeneity. For a one-way design, the assumption states that the variance of Y is the same across all conditions. In other words, we assume that the within-group variance of Y is the same for all levels of the factor.

This assumption extends to factorial designs. More specifically, the assumption states the within-group variance is the same across all experimental conditions. In this case it means that the variance is the same for all six conditions. Again, we can test this assumption! Change the syntax as follows:

 

UNIANOVA Tip BY Behavior Gender

     /METHOD=SSTYPE(3)

     /INTERCEPT=INCLUDE

     /PRINT=DESCRIPTIVE HOMOGENEITY

     /CRITERIA=ALPHA(.05)

     /DESIGN=Behavior Gender Behavior*Gender.

 

Run the syntax!

 


SPSS now includes the Levene’s test.

As you know, the Levene’s test tests H0: “equal variances” against H1: “non-equal variances”.

Is the assumption met?

Correct answer
  Yes 
  No 
General answer comments
The Levene’s test is not significant, F(5,54) = 1.644, p = .164.

Hence, we don’t have evidence that this assumption is violated.

 


Why is it important to evaluate the assumption?
Or, in other words, what goes wrong if the assumption is grossly violated?
General answer comments
If this assumption is grossly violated (and the sample sizes of the groups are different), we can no longer rely on the p-values when drawing conclusions about significance.


## Do it yourself!


In the previous assignments we helped you carrying out a factorial ANOVA step by step. In this assignment you need to carry out the analysis independently.

Consider the data on the effects of the behavior of the hiking guide on feelings again (we used these data last week too). Open the datafile hiking.sav Download hiking.sav. The data file also contains data on weather. The weather conditions may be a moderator variable.

Let’s run a factorial ANOVA in which you include weather as second factor. So, you need to make a model with weather and behavior as independent variables, and feelings as the dependent variable.

Make sure that you:

Run the ANOVA
Ask for Levene’s test
Ask for a plot
 


Is the assumption of homogeneity reasonable with an alpha of .05?
Correct answer
  No 
  Yes 
 


If done correctly, you should’ve concluded that the assumption of homogeneity is violated.

However, this is not much of a problem if the sample sizes of the groups are (roughly) equal to one another. Fortunately, in this case, all groups are of equal size, so we can safely carry on with the analysis.

What is the explained variance of the main effects and interaction effects together?

Correct Answers
0.239 (with margin: 0.005)
 


Is the explained variance of the whole model significant? (use α = .05)
Correct answer
  Yes 
  No 
 


Is the interaction effect significant when tested at the 5% level?
Correct answer
  Yes 
  No 
General answer comments
The interaction effect is significant when using α=.05, F(4,90) = 2.547, p = 0.045.

Hence, the next step is to do a test of simple effects.

 


If done correctly, you should’ve concluded that the interaction effect is significant.

Run the appropriate analysis that allows you to describe the trends (i.e. what the effects look like).
Hint: request a plot from SPSS.

Describe the trends in your own words.

General answer comments
We see a significant effect of behavior when the weather was good, F(4,90) = 3.864, p = .006, while the effect of behavior is not significant when the weather was bad, F(4,90) = 1.320, p = .269.

The lines in the plot are not parallel, also pointing to an interaction effect.

 


What is your final conclusion?

(Hint: Inspect the significance tests of the simple effects analysis.)

General answer comments
Results suggest that when the weather is good, joking and singing significantly improves the participants’ feelings about the guide.

When the weather is bad, the behavior of the guide does not have much influence.

 
