---
title: "Welcome to 'Statistics 1'"
author:   "Caspar J. van Lissa"
date:     "`r format(Sys.Date(), '%d %b %Y')`"
format: revealjs
---

```{r}
library(kableExtra)
library(tidySEM)
options(knitr.kable.NA = '')
set.seed(1)
```


```{r}
#| label = "setup",
#| include = FALSE
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
source("functions.R")
```

## Today

* Course overview
* Examination
* Using course resources
* Expectations
* Motivation

## Course Overview

Forget about specific techniques; you will learn:


* Essential skill: extract insight from data
* “data literacy”: learn how to use quantitative data to make evidence-based decisions
* Basic statistics and data analysis
* To independently analyze data
* Interpret and report your findings
* Assess analyses performed by others
    
## Course Structure


* ~14 lectures
* ~14 lab sessions

## Course Philosophy

* Flipped classroom
* Challenge-based Learning
* Group work
* Spread-out assessment, lots of formative assessment

## What we do in the Lectures

* Practice exam questions
* Open-ended questions to guide class discussion
* Q&A about conceptual questions (i.e., "getting it")
* Specific conceptual questions about your portfolio assignment

## What we do in the Tutorials

* Working through exercises to build competence in data analysis  (optionally: with your group members)
* Working on portfolio assignments (with your group members)
* Q&A about applied statistics (i.e., "doing it")
* Specific applied questions about portfolio

## What you do at Home

* Read the GitBook
* Study the lecture video
* Make formative quiz to assess your level of understanding
* Meet regularly with portfolio group to study together

## Examination

* 1 multiple choice exam
    + Split into 2 sessions
    + 60% of grade
* 1 portfolio, split into 2 group assignments
    + 40% of grade
  
## Course Resources

The GitBook contains **all** relevant information

Let me show you...

## Expectations

I assume that you...

* Take responsibility for your learning
* Read the Syllabus (which is also the first page of the GitBook)
* Show up on time to all sessions
* Have studied the lecture video
* Participate actively during lectures and lab sessions
* Co-create supportive, safe, fun, empathic space for learning
    + Learning meansbeing challenged. Challenges come in many forms
    
## Hierarchy for Questions

If you have questions:

1. Check the GitBook
2. Ask your groupmate
3. Ask on the Discussion page
4. Only if your question is still not answered, then contact me

# Motivation

<!-- ## I don't Want to be Here -->

<!-- * I would rather be climbing mountains or flying small planes -->
<!-- * Why am I here? -->
<!-- * Because one of my core values is contribution, and I can contribute by teaching you -->
<!-- * Because one of my core values is personal growth, and I can grow as a teacher by continually improving this course -->
<!-- * Because one of my core values is connection, and I can connect with you and have a "high-quality learning experience" together -->

<!-- What about you? -->


## Set Your Standards

Forget about passing. What do **YOU** want to get out of this course?

* Define your expectations for your performance in this course
* Set standard for grades, understanding, teamwork, etc
* Aim higher to truly grasp the material
* Become a critical thinker

## Make a Decision

**De-cidere: Cutting off alternative paths**

Are you ready to become really GREAT at statistics?

Deciding means committing to a specific result, and closing the door to any other outcome.

Decide what kind of student you will be.

* Commit to your standard
* Be present
* Complete (group) assignments on time
* Seek help when needed:
		1. Course manual
		2. Groupmates
		3. Discussion page
		4. Teacher
* Minimal effort is not an option if you want to succeed

## Really believe it

Complete trust in your ability to pass this course

I had a 3.4 in high school math - now I am a statistics professor

The question is not: can I do this? The question is: How do I do this!

## Decisions Determine Outcomes

What percentage is aptitude, and what percentage is elbow grease?

In my case, probably 10% aptitude, 90% hard work

What about you?

Whatever you have control over - focus on that

* Make decisions to study, practice, and engage with the material
* Shape your reality

## Motivation

What are your values?

* Connection
* Growth
* Security
* Adventure
* Contribution

Each of these can bring pain or pleasure

## Motivation 2

**Pain associated with studying**

* Adventure/connection: Takes time you could have used for fun
* Growth/security: Is challenging
* Growth/security: You might fail sometimes
* Adventure: Can be boring

## Motivation 3

**Pain associated with NOT studying**

* Growth/security: Fail the course, have to go through all that again
* Connection/growth/security: Feel unprepared, embarrassed, judged, etc.
* Connection: Fall short of agreements with group mates
* Adventure: Not able to go on exchange

## Motivation 4

**Pleasure associated with studying**

* Connection: Work with group members, make new friends
* Growth: Develop valuable real-life skills in data analysis
* Contribution: Be a valuable team member
* Security: Feel secure that you won't have to extend your studies
* Adventure: Time invested pays off later; reward yourself with "adventure" (whatever it means to you)

## Increasing Motivation

* Make the pain associated with NOT studying bigger, and increase reward associated with studying
* Reward yourself regularly! (celebrate every small success, Caryl Rusbult)
* Create accountability: share your goals with team members
* Agree on rewards: make it fun (social), create a point system, give each other compliments, whatever works for you
* Or agree on "punishments": lowest grade pays a round of drinks, etc.


## Reinforce Good Behaviors

**Behavior that is consistently reinforced will become a habit**

* Create positive study habits: 
  + Just commit to 5 minutes, every morning (5 soon turns into 50)
	+ Schedule regular study times (put it in your agenda)
	+ Reward small successes (do a bit of work before exercise, or before having a nice snack, or before going out with a friend)
* _Habits_ will serve you throughout your academic career and life


## Problems Are Not Permanent

**Struggle in statistics?**

* Challenges are temporary hurdles
* With persistence, you can grok (understand) any topic.
* Growth mindset: Challenges are also opportunities to grow


## Ask the Right Questions

* Not "is this part of the exam"?
* "How does this apply to my academic and real-world interests?"
* "Why does this matter?"
* "What is my fellow student - who understands the material - doing differently than me?"
* Curiosity drives learning: The more you question, the more you learn.

## Handle Your Emotions

* S#*$ gets personal!
* Emotions are part of the learning and teamwork process
* Remember a time when you felt a similar emotion and realize that you've successfully handled this before
* These are also learning opportunities
* Expect emotions and prepare strategies to cope
  + Frustration: Take a breather, take smaller steps, ask for help
  + In team work: use co-regulation (e.g., empathy and constructive problem solving) to resolve negative emotions together


**Become Resilient:** Your ability to manage stress and emotions will contribute to your success in this course and life

## Group Work

* Establish shared goals
* Write and sign a "Teamwork Contract"
* Communicate Your Expectations
		+ Don't expect people to live up to standards you never clearly communicated
* Create roles (either for the whole assignment, or rotate every week)
		+ Planner, timekeeper, proofreader, counter-thinker, coach, social lubricator
* Check in with each other regularly (how's it going with you and with our project)
* Reflect at the end (this is part of the grade)


## Urgency vs. Importance

* It is possible to work hard and still fall short (work smart, not hard)
* Focus on what's important:

		+ Studying: Understanding > passing exam questions
		+ Assignments: Good/clear > long and fluffy
		+ Scope: Material offered > other/extra material
		+ Group work: Teamwork > individual performance


## Journey to Mastery

* Yes, statistics is a valuable skill
* This is not just about passing a course: It's about becoming the kind of person who successfully navigates challenges and grows
* Embrace the journey: Use these principles to guide you through your studies

## Exercises

1. Write down (at least) 3 things **YOU** want to get out of this course
2. Write down 3 decisions you are ready to make in relation to this course
3. Pick your top 3 values (Connection, Growth, Security, Adventure, Contribution, or something else?)
4. Think of 1 (funny OR serious) way to increase pain associated with not studying
5. Think of 1 way to increase pleasure associated with studying
6. Write down one intention for your group work this semester

